Parameter Recovery Validation

### 2025-05-28 nbs/concepts/parameter-recovery-validation/index.qmd
docs: clarify interpretation of expression phases and singular point
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@

This piecewise framework enables representation of diverse gene expression patterns:

1. **Activation patterns**: Low $\alpha^*_{\text{off}i}$, early $t^*_{0,\text{on}i}$, high $\alpha^*_{\text{on}i}$
2. **Decay-only patterns**: High $\alpha^*_{\text{off}i}$, $t^*_{0,\text{on}i} > T^*_M$ (onset beyond observation window)
3. **Transient activation**: Early $t^*_{0,\text{on}i}$, short $\delta^*_i$, moderate $\alpha^*_{\text{on}i}$
4. **Sustained activation**: Early $t^*_{0,\text{on}i}$, long $\delta^*_i$, high $\alpha^*_{\text{on}i}$

>>>>

**Special case interpretation ($\gamma^*_i = 1$):**

The case $\gamma^*_i = 1$ represents **balanced splicing-degradation kinetics** where mRNA degradation occurs at exactly the same rate as pre-mRNA splicing. This creates a mathematical resonance that fundamentally changes the system's dynamics:

- **Biological meaning**: The characteristic timescales for splicing ($1/\beta_i$) and degradation ($1/\gamma_i$) are identical
- **Mathematical consequence**: The standard exponential solutions become degenerate, requiring the additional $\tau_{\text{on}} e^{-\tau_{\text{on}}}$ term
- **Physical interpretation**: The $\tau e^{-\tau}$ term represents the accumulation of mRNA over time when production and degradation rates are perfectly balanced
- **Parameter recovery implications**: Inference near $\gamma^*_i = 1$ may exhibit slower convergence due to the discontinuous derivative in the analytical solutions
- **Numerical considerations**: The transition between $\gamma^*_i \neq 1$ and $\gamma^*_i = 1$ solutions requires careful handling to maintain numerical stability

@@

This piecewise framework enables representation of diverse gene expression patterns through specific parameter combinations. Each pattern is defined by quantitative parameter ranges that create distinct temporal dynamics:

#### **1. Activation Patterns**
**Characteristics**: Genes that start with low expression and become strongly activated during the process.

**Parameter specifications**:
- $\alpha^*_{\text{off}i} < 0.2$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.5$ (strong activation)
- $t^*_{0,\text{on}i} < 0.4$ (early activation onset)
- $\delta^*_i > 0.3$ (sustained activation duration)
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 7.5$ (strong upregulation)

**Biological examples**: Developmental transcription factors, cell cycle genes, stress response genes

#### **2. Decay-Only Patterns**
**Characteristics**: Genes that start with high expression and gradually decrease without reactivation.

**Parameter specifications**:
- $\alpha^*_{\text{off}i} > 0.5$ (high basal transcription)
- $t^*_{0,\text{on}i} > T^*_M$ (activation onset beyond observation window)
- $\delta^*_i$ (irrelevant, no activation observed)
- **Effective dynamics**: Monotonic decay from high initial levels

**Biological examples**: Maternal transcripts during development, housekeeping genes during stress, pluripotency factors during differentiation

#### **3. Transient Activation Patterns**
**Characteristics**: Genes that show brief, pulse-like activation followed by return to baseline.

**Parameter specifications**:
- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.0$ (moderate to strong activation)
- $t^*_{0,\text{on}i} < 0.5$ (early to mid-process activation)
- $\delta^*_i < 0.3$ (brief activation duration)
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (clear but not extreme upregulation)

**Biological examples**: Immediate early genes, signaling molecules, transient developmental regulators

#### **4. Sustained Activation Patterns**
**Characteristics**: Genes that become activated and remain highly expressed throughout most of the process.

**Parameter specifications**:
- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.0$ (strong activation)
- $t^*_{0,\text{on}i} < 0.3$ (early activation onset)
- $\delta^*_i > 0.6$ (long activation duration)
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (sustained upregulation)

**Biological examples**: Lineage-specific transcription factors, structural proteins, metabolic enzymes

#### **Pattern Classification Criteria**

For parameter recovery validation, genes can be automatically classified using these decision rules:

```python
def classify_gene_pattern(alpha_off, alpha_on, t_on, delta, T_max):
    """Classify gene expression pattern based on inferred parameters."""
    fold_change = alpha_on / alpha_off

    if t_on > T_max:
        return "decay_only"
    elif alpha_off < 0.2 and alpha_on > 1.5 and t_on < 0.4 and delta > 0.3:
        return "activation"
    elif alpha_off < 0.3 and fold_change > 3.3 and t_on < 0.5 and delta < 0.3:
        return "transient"
    elif alpha_off < 0.3 and fold_change > 3.3 and t_on < 0.3 and delta > 0.6:
        return "sustained"
    else:
        return "intermediate"  # Mixed or boundary cases
```

#### **Validation Strategy for Pattern Recovery**

Parameter recovery validation should test the model's ability to:

1. **Correctly classify known patterns**: Generate synthetic data with known pattern types and verify classification accuracy
2. **Recover pattern-specific parameters**: Ensure fold-changes, timing, and duration parameters are accurately inferred
3. **Distinguish similar patterns**: Separate transient vs. sustained activation, activation vs. decay-only
4. **Handle boundary cases**: Genes with intermediate parameter values that don't clearly fit one pattern
